\ttitle

### 2015-08-12 Chapters/Introduction.tex
Discussion - 2/3 complete
- Minor changes in introduction
- Fixed effing error in zot.bib as a consequence of error in junejo abstract with Tex commands embedded - WHY???
- Fixed a few minor issues with main.tex
Introduction - revision to section 1.3
- Spelling, grammar, style corrections

\section{Pancreatic Neoplasia}

@@

Pancreatic ductal adenocarcinoma is the most common cancer of the pancreas. However, the head of the pancreas is anatomically related to several other epithelium lined structures that can also give rise to cancers. These include the distal common bile duct that can give rise to cholangiocarcinoma, the duodenum that can give rise to duodenal adenocarcinoma and the ampulla that can give rise to ampullary adenocarcinoma. The endocrine portion of the pancreas can give rise to a variety of tumours that are collectively called neuroendocrine tumours (NET). The milieu of tumours is complicated by other neoplasia such as intra-ductal papillary neoplasms (IPMN) as well as rare stromal tumours. Occasionally, chronic pancreatitis may present with features similar to pancreatic cancer and can be morphologically, radiologically and histologically difficult to differentiate from cancer.

Pancreatic cancer is the tenth most common cancer in the UK but the fifth most common cause of cancer death with only 21\% surviving beyond the first year and 3\% surviving beyond 5 years.\parencite{cancerresearchuk_cancer_2014} The majority of patients (80-85\%) with pancreatic cancer present with inoperable disease.\parencite{cancerresearchuk_cancer_2014,sener_pancreatic_1999}

In patients with resectable disease, surgery \parencite{sener_pancreatic_1999, sohn_resected_2000,geer_prognostic_1993} followed by adjuvant chemotherapy \parencite{neoptolemos_randomized_2004, neoptolemos_adjuvant_2009} remains the primary modality of cure. However, major pancreatic surgery places significant physiological stresses on multiple organ systems. The ability of the cardiac and respiratory systems, in particular, to cope with the increased physiological demand placed by general anaesthesia and major pancreatic surgery plays an important role in determining outcome after surgery.

@@

The anatomical location of the pancreas, deep within the retroperitoneum surrounded by numerous vital blood vessels including the coeliac trunk and its branches, the superior mesenteric artery, portal vein and superior mesenteric vein as well as proximity to other viscera such as the stomach, duodenum, transverse colon result in early involvement of these structures even by relatively small tumours. Moreover, symptoms are often absent in the early stages and when present are too non-specific to help with diagnosis. Obstructive jaundice is the most common presenting symptom and painless, obstructive jaundice in an elderly patient should always raise the suspicion of a neoplastic process in the head of the pancreas or the periampullary region. Other non-specific symptoms include weight loss, early satiety, vomiting, fatigue and pain in the epigastrium or the back.

@@

Aside from a thorough history, clinical examination, blood tests including liver function tests, diagnosis requires cross-sectional imaging in the form of a contrast-enhanced computerised tomogram (CECT) of the abdomen using a pancreas-specific protocol (a modified form of the portal-venous phase). CECT of the pancreas when combined with CT Thorax also provides accurate information on staging of the disease with regards to metastasis and this can be supplemented by further imaging such as Positron Emission Tomography (PET-CT) or contrast-enhanced MRI Liver in specific cases. CECT-pancreas is also useful for assessing local resectability with regards to vascular involvement. Endoscopic ultrasound (EUS) is also useful in assessing vascular involvement and for obtaining tissue samples for histological examination. In jaundiced patients, endoscopic retrograde cholangio pancreatography (ERCP) plays an important role in the alleviation of jaundice by placing stents across the obstructed bile ducts, accurate visualisation of the biliary anatomy as well as obtaining brushings from within the bile ducts for cytological examination. The role of preoperative biliary drainage is discussed in more detail in Section \ref{sec:preoperative_biliary_drainage}.

\subsection{Treatment of pancreatic cancer}
Pancreaticoduodenectomy followed by adjuvant chemotherapy offers the only chance of cure in patients with resectable pancreatic cancer who are fit enough to undergo surgery. In patients with unresectable disease or who are not fit to undergo surgery, palliative chemotherapy plays a limited role in prolonging survival. Assessing the resectability is discussed in the next section while the assessment of patient fitness and the impact of comorbidity are discussed in detail in section \ref{sec:comorbidity_risk_stratification} on p\pageref{sec:comorbidity_risk_stratification}.

\section{Surgical treatment of pancreatic cancer}
Pancreaticoduodenectomy remains a technically challenging and complex surgical procedure over a hundred years after its description. The procedure was performed as a two-stage operation by a German surgeon, Walther Kausch in 1909 at Augusta-Viktoria-Krankenhaus in Berlin-Schöneberg.\parencite{kausch_carcinom_1912}. The operation was further popularised initially as a two-stage procedure by Whipple\parencite{whipple_treatment_1935} before evolving into the current single stage operation by the 1950s.\parencite{whipple_rationale_1941,whipple_radical_1950}

\subsection{Patient selection}
\subsubsection{Resectability criteria}
Resectable pancreatic cancer is defined as a tumour that
- does not involve the coeliac axis or the superior mesenteric artery
- and is not associated with distant metastatic disease

Tumours involving the portal vein or superior mesenteric vein are considered borderline resectable and can still be resected completely (R0) with en-bloc venous resection in selected cases. The role of neoadjuvant therapy such as chemotherapy with or without local radiotherapy and newer treatment modalities such as electroporation  and intra-peritoneal chemotherapy is the subject of ongoing clinical research in patients with locally advanced disease and in some patients with metastatic disease. Early reports suggest that some of these patients can undergo potentially curative surgery after such neo-adjuvant treatment.

\subsubsection{Patient factors}

@@

After a comprehensive preoperative work-up including both assessments of the tumour as well as patient fitness, informed consent was obtained. Patients received thrombo-prophylaxis on the night before surgery which was continued until discharge from hospital. General anaesthesia with complete muscle relaxation was used in all patients. Epidural analgesia was used routinely in patients during the early part of the study period while all patients in the later half of the study period received spinal diamorphine. Antibiotic prophylaxis is administered at induction. While the use of Octreotide, a somatostatin analogue, to reduce the incidence of postoperative pancreatic fistula is still under debate, it was routinely used in all patients at this centre. Octreotide was administered intra-operatively (200 micrograms subcutaneous injection at induction and repeated after 4 hours) and was continued for 5 days postoperatively (200 micrograms subcutaneous injection administered three times a day).

@@

\subsection{Postoperative care}
All patients were routinely admitted to the Surgical High Dependency Unit unless intra-operative events necessitated admission to the Intensive Care Unit. A standardised regimen of intravenous fluids, naso-jejunal feeding, mobilisation and physiotherapy was implemented in all patients. Standard physiological parameters including haemodynamic parameters, renal function and arterial blood gases were used to monitor adequate end organ perfusion. All patients received proton pump inhibitors and Octreotide Patients were discharged to the general surgical ward as early as possible.

@@

Postoperative pancreatic fistula is one of the most dreaded complications after a pancreaticoduodenectomy and can be associated with significant short-term morbidity as well as long-term disability. The reported incidence of postoperative pancreatic fistula varies from 2\% to 30\% after pancreaticoduodenectomy.\parencite{yeo_six_1997,deoliveira_assessment_2006,bassi_postoperative_2005,winter_biochemical_2007,pratt_risk_2008} 
The variation in reported incidence has been largely due to lack of clear definition of what constituted a postoperative pancreatic fistula. It can be a result of breakdown or poor healing at the pancreatico-jejunostomy/pancreatico-gastrostomy or may be the result of direct parenchymal leak unrelated to the anastomosis. It is now generally accepted that 1 in 4 patients will develop a pancreatic fistula as defined by the International Study Group for Pancreatic Fistula (ISGPF) which has published a consensus statement on the definition and grading of postoperative pancreatic fistula.\parencite{bassi_postoperative_2005} A postoperative pancreatic fistula is defined as drain output of any measurable quantity after the third postoperative day with amylase content greater than three times the upper limit of the normal serum amylase value at the laboratory used for testing. Three grades of postoperative pancreatic fistula have been defined based on clinical severity as described in Table \ref{table:isgps_popf} on p\pageref{table:isgps_popf}. Grade B and C fistulae are considered to be clinically significant as they alter patient management and are often associated with other secondary complications such as intra-abdominal sepsis, post-pancreatectomy haemorrhage or delayed gastric emptying. They may require further intervention (either radiological or operative) and may result in prolonged critical care support.

@@

Haemorrhage is often secondary to non-healing of the pancreatico-jejunal anastomosis leading to leakage of amylase-rich pancreatic juices into the retroperitoneum or secondary to intra-abdominal sepsis or bile leak.\parencite{tien_risk_2005, koukoutsis_haemorrhage_2006, choi_delayed_2004, balladur_bleeding_1996} This can then lead to erosion of ligated blood vessels, most commonly the stump of the gastro-duodenal artery. Post-pancreatectomy haemorrhage is often managed with angiographic embolisation of the bleeding vessel and surgical intervention is only rarely required. The grading of severity of post-pancreatectomy haemorrhage as described by the International Study Group of Pancreatic Surgery\parencite{wente_postpancreatectomy_2007} is shown in Table \ref{table:isgps_pph} on p\pageref{table:isgps_pph}.

\subsubsection{Clavien-Dindo classification of complications}
\input{Tables/c1_clavien_dindo}
A number of other adverse events may occur following pancreaticoduodenectomy including cardiopulmonary complications such as myocardial infarction, cardiac arrhythmias, pneumonia and pleural effusions, wound complications such as wound sepsis and dehiscence, intra-abdominal problems including intra-abdominal sepsis, leakage from the hepatico-jejunostomy or the gastro-jejunostomy, infected fluid collections, renal dysfunction, deep vein thrombosis, pulmonary embolism, etc. The Clavien-Dindo system grades the severity of complications based on the impact the complication has on the management of the patient and has been validated on large numbers of surgical patients.\parencite{clavien_clavien-dindo_2009, dindo_classification_2004} This is summarised in Table \ref{table:clavien-dindo} on p\pageref{table:clavien-dindo} and has been used to grade complications in this thesis. Complications of grade 3 or greater are considered clinically significant as they require further procedures either under local or general anaesthesia, single or multiple organ support or result in death while grades 1 and 2 are managed with ward-based care including antibiotics, wound management, supplemental enteral or parenteral nutrition and similar interventions.

@@

While surgery is the primary modality of curative treatment in patients with resectable pancreatic cancer, there is clear evidence that adjuvant chemotherapy improves survival and is now the standard of care. The European Study Group for Pancreatic Cancer (ESPAC) have conducted some of the largest randomised controlled trials of adjuvant chemotherapy and chemoradiotherapy versus observation after potentially curative surgery and have reported that chemotherapy prolonged survival after surgery while chemoradiotherapy had a harmful effect. \parencite{neoptolemos_adjuvant_2001, neoptolemos_randomized_2004, neoptolemos_adjuvant_2009, neoptolemos_adjuvant_2010}

The ESPAC-1 trial compared the use of chemotherapy with Fluorouracil and folinic acid or chemoradiotherapy with 20Gy radiotherapy and 5-Fluorouracil (5-FU) versus observation only in patients who underwent potentially curative surgery for pancreatic ductal adenocarcinoma. The 5-year survival rate in patients who received chemotherapy was 20\% while it was 8\% in patients who did not receive demonstrating a clear survival benefit for adjuvant chemotherapy. However, adjuvant chemoradiotherapy resulted in 5-year survival dropping from 20\% to 10\% and has largely been abandoned in the adjuvant setting. The ESPAC-3 trial compared the efficacy of 5-FU and Folinic acid with Gemcitabine and reported that Gemcitabine was not superior to 5-FU/Folinic acid.

However, 25\% to 50\% of patients who undergo surgery for pancreatic ductal adenocarcinoma end up either not receiving or not completing adjuvant chemotherapy. \parencite{sohn_resected_2000, neoptolemos_randomized_2004, oettle_adjuvant_2007} A number of factors have been identified for the non-receipt of adjuvant therapy including comorbdity, postoperative complications and favourable pathological features.\parencite{spitz_preoperative_1997, aloia_delayed_2007, oettle_adjuvant_2007, merchant_adjuvant_2009, russ_impact_2009}

@@

More recently, cross-sectional imaging including computed tomography (CT) and magnetic resonance imaging (MRI) have become widely available. Nearly all patients with gastrointestinal cancer undergo either CT or MRI or both before commencing treatment. Accurate measurements of subcutaneous adipose tissue, visceral adipose tissue and skeletal muscle can be made relatively easy either from the entire series of images or estimates can be made from measurements made at certain predetermined anatomical levels. Both specialised software or readily available generic image analysis tools have been used to make these measurements.

The proportion of the total cross-sectional area that is made of adipose tissue or skeletal muscle at the level of the 3$^{rd}$  or the 4$^{th}$ lumbar vertebra is commonly used as a surrogate for total body measurements. Thickness of retro-renal visceral fat at the level of the renal vein has also been reported to be usable as a surrogate for total visceral adiposity.

>>>>

\section{Pancreatic Neoplasia}
%This is the only paper I need for this entire section - pancreatic_section_of_the_british_society_of_gastroenterology_pancreatic_society_of_great_britain_and_ireland_association_of_upper_gastrointestinal_surgeons_of_great_britain_and_ireland_royal_college_of_pathologists_special_interest_group_for_gastro-i_guidelines_2005

@@

Pancreatic ductal adenocarcinoma is the most common cancer of the pancreas. However, the head of the pancreas is anatomically related to several other epithelium lined structures that can also give rise to cancers. These include the distal common bile duct (cholangiocarcinoma), the duodenum (duodenal adenocarcinoma) and the ampulla (ampullary adenocarcinoma). Neuroendocrine tumours (NET) are a collection of a variety of neoplasia that originate from the endocrine portion of the pancreas. This milieu is complicated by other neoplasia such as intra-ductal papillary neoplasms (IPMN) as well as rare stromal tumours. Occasionally, chronic pancreatitis may present with features similar to pancreatic cancer and can be morphologically, radiologically and histologically difficult to differentiate from cancer.

Pancreatic cancer is the tenth most common cancer in the UK but the fifth most common cause of cancer death. Only 21\% survive beyond the first year and 5-year survival is 3\%. \parencite{cancerresearchuk_cancer_2014} The majority of patients (80-85\%) with pancreatic cancer present with inoperable disease.\parencite{cancerresearchuk_cancer_2014,sener_pancreatic_1999}

In patients with resectable disease, surgery \parencite{sener_pancreatic_1999, sohn_resected_2000,geer_prognostic_1993} followed by adjuvant chemotherapy \parencite{neoptolemos_randomized_2004, neoptolemos_adjuvant_2009} offers the only chance of a cure. However, major pancreatic surgery places significant physiological stresses on multiple organ systems. The ability of the cardiac and respiratory systems in particular, to cope with the increased physiological demand placed by general anaesthesia and major pancreatic surgery plays an important role in determining outcome after surgery.

@@

The anatomical location of the pancreas, deep within the retroperitoneum, surrounded by numerous vital blood vessels including the coeliac trunk and its branches, the superior mesenteric artery, portal vein and superior mesenteric vein as well as proximity to other viscera such as the stomach, duodenum, transverse colon result in early involvement of these structures even by relatively small tumours. Moreover, symptoms are often absent in the early stages and when present are too non-specific to help with diagnosis. Obstructive jaundice is the most common presenting symptom and painless, obstructive jaundice in an elderly patient should always raise the suspicion of a neoplastic process in the head of the pancreas or the periampullary region. Other non-specific symptoms include weight loss, early satiety, vomiting, fatigue and pain in the epigastrium or the back. Occasionally, the first presentation of a pancreatic neoplasm may be in the form of an episode of acute pancreatitis.

@@

Initial assessment includes a thorough history, clinical examination and blood tests including full blood count, renal function tests, liver function tests and tumour markers, especially CA 19-9. However, definitive diagnosis requires cross-sectional imaging in the form of a contrast-enhanced computerised tomogram (CECT) of the abdomen using a pancreas-specific protocol (a modified form of the portal-venous phase). CECT of the pancreas when combined with CT Thorax also provides accurate information on staging of the disease with regards to metastasis and this can be supplemented by further imaging such as Positron Emission Tomography (PET-CT) or contrast-enhanced Magnetic Resonance Imaging (MRI) of the liver in specific cases. CECT-pancreas is also useful for assessing local resectability with accurate information regarding vascular involvement. Endoscopic Ultrasound (EUS) is used in selected patients to assess vascular involvement and for obtaining tissue samples for histological examination. In jaundiced patients, endoscopic retrograde cholangio pancreatography (ERCP) plays an important role in the alleviation of jaundice by placing stents across the obstructed bile duct, accurate visualisation of the biliary anatomy as well as obtaining brushings from within the bile duct for cytological examination. The role of preoperative biliary drainage is discussed in more detail in Section \ref{sec:preoperative_biliary_drainage}.

\subsection{Treatment of pancreatic cancer}
Pancreaticoduodenectomy followed by adjuvant chemotherapy offers the only chance of cure in patients with resectable pancreatic cancer who are fit enough to undergo surgery. In patients with unresectable disease or who are not fit to undergo surgery, palliative chemotherapy plays a limited role in prolonging survival. Assessing the resectability of lesions of the head of the pancreas is discussed in the next section while the assessment of patient fitness and the impact of comorbidity are discussed in detail in section \ref{sec:comorbidity_risk_stratification} on p\pageref{sec:comorbidity_risk_stratification}.

\section{Surgical treatment of pancreatic cancer}
Pancreaticoduodenectomy remains a technically challenging and complex surgical procedure over a hundred years after its description. Excision of cancer of the head of the pancreas was first described as a two-stage operation by a German surgeon, Walther Kausch in 1909 at Augusta-Viktoria-Krankenhaus in Berlin-Schöneberg.\parencite{kausch_carcinom_1912}. The operation was further popularised initially as a two-stage procedure by Whipple\parencite{whipple_treatment_1935} before evolving into the current single stage operation by the 1950s.\parencite{whipple_rationale_1941,whipple_radical_1950}

\subsection{Patient selection}
\subsubsection{Resectability criteria}
Resectable pancreatic cancer is defined as a tumour that does not involve the coeliac axis or the superior mesenteric artery and is not associated with distant metastatic disease. Tumours involving the portal vein or superior mesenteric vein are considered borderline resectable and can still be resected completely (R0) with en-bloc venous resection in selected cases.\parencite{fuhrman_rationale_1996} The role of neoadjuvant therapy such as chemotherapy with or without local radiotherapy and newer treatment modalities such as electroporation\parencite{bower_irreversible_2011}  and intra-peritoneal chemotherapy is the subject of ongoing clinical research in patients with locally advanced and/or metastatic disease. Early reports suggest that some of these patients can undergo potentially curative surgery after neo-adjuvant treatment.

%10/08/2015 - Review completete to here
\subsubsection{Patient factors}

@@

A comprehensive preoperative evaluation was performed including staging of the disease for local resectability and absence of metatstatic disease, assessment of patient fitness and anaesthetic review.  All patients with suspected cancer were discussed at the pancreato-biliary multidisciplinary team meeting. A fully informed consent was obtained after discussion regarding treatment options, the details of a pancreaticoduodenectomy, expected perioperative course, potential risks and follow-up care.

Patients received thrombo-prophylaxis on the night before surgery which was continued until discharge from hospital. General anaesthesia with complete muscle relaxation was used in all patients. Epidural analgesia was used routinely in patients during the early part of the study period while all patients in the later half of the study period received spinal diamorphine. Prophylactic antibiotics were administered at induction. While the use of Octreotide, a somatostatin analogue, to reduce the incidence of postoperative pancreatic fistula is still under debate, it was routinely used in all patients at this centre. Octreotide was administered intra-operatively (200 micrograms subcutaneous injection at induction and repeated after 4 hours) and was continued for 5 days postoperatively (200 micrograms subcutaneous injection administered three times a day).

@@

\subsection{Postoperative care}
All patients were routinely admitted to the Surgical High Dependency Unit unless intra-operative events necessitated admission to the Intensive Care Unit. A standardised regimen of intravenous fluids, naso-jejunal feeding, mobilisation and physiotherapy was implemented in all patients. Standard physiological parameters including haemodynamic parameters, renal function and arterial blood gases were used to monitor adequate end organ perfusion. All patients received proton pump inhibitors and Octreotide. Patients were discharged to the general surgical ward as early as possible. Amylase content in the drain fluid was measured on the third post-opertive day to aid in the diagnosis of post-operative pancreatic fistula. Surgical drains and naso-gastric/naso-jeujunal tubes were removed at surgeon's discretion. Patients were discharged when medically fit.

Histology of the removed specimen was discussed at the multi-disciplinary team meeting in cancer patients and suitable patients were referred for adjuvant chemotherapy.

@@

Postoperative pancreatic fistula is one of the most dreaded complications after a pancreaticoduodenectomy and can be associated with significant short-term morbidity as well as long-term disability. The reported incidence of postoperative pancreatic fistula varies from 2\% to 30\% after pancreaticoduodenectomy. \parencite{yeo_six_1997, deoliveira_assessment_2006, bassi_postoperative_2005, winter_biochemical_2007, pratt_risk_2008} The variation in reported incidence has been largely due to lack of clear definition of what constituted a postoperative pancreatic fistula. It can be the result of breakdown or poor healing at the pancreatico-jejunostomy/ pancreatico-gastrostomy or may be the result of direct parenchymal leak unrelated to the anastomosis. It is now generally accepted that 1 in 4 patients will develop a pancreatic fistula as defined by the International Study Group for Pancreatic Fistula (ISGPF) which has published a consensus statement on the definition and grading of postoperative pancreatic fistula.\parencite{bassi_postoperative_2005} A postoperative pancreatic fistula is defined as drain output of any measurable quantity after the third postoperative day with amylase content greater than three times the upper limit of the normal serum amylase value at the laboratory used for testing. Three grades of postoperative pancreatic fistula have been defined based on clinical severity as described in Table \ref{table:isgps_popf} on p\pageref{table:isgps_popf}. Grade B and C fistulae are considered to be clinically significant as they alter patient management and are often associated with other secondary complications such as intra-abdominal sepsis, post-pancreatectomy haemorrhage or delayed gastric emptying. They may require further intervention (either radiological or operative) or prolonged critical care support.

@@

Haemorrhage is often secondary to non-healing of the pancreatico-jejunal anastomosis leading to leakage of amylase-rich pancreatic juices into the retroperitoneum or secondary to intra-abdominal sepsis or bile leak.\parencite{tien_risk_2005, koukoutsis_haemorrhage_2006, choi_delayed_2004, balladur_bleeding_1996} This can then lead to erosion of ligated blood vessels, most commonly the stump of the gastro-duodenal artery. Post-pancreatectomy haemorrhage is usually managed with angiographic embolisation of the bleeding vessel and surgical intervention is only rarely required. The grading of severity of post-pancreatectomy haemorrhage as described by the International Study Group of Pancreatic Surgery\parencite{wente_postpancreatectomy_2007} is shown in Table \ref{table:isgps_pph} on p\pageref{table:isgps_pph}.

\subsubsection{Clavien-Dindo classification of complications}
\input{Tables/c1_clavien_dindo}
A number of other adverse events may occur following pancreaticoduodenectomy including cardiopulmonary complications such as myocardial infarction, cardiac arrhythmias, pneumonia and pleural effusions, wound complications such as wound sepsis and dehiscence, intra-abdominal problems including intra-abdominal sepsis, leakage from the hepatico-jejunostomy or the gastro-jejunostomy, infected fluid collections, renal dysfunction, deep vein thrombosis, pulmonary embolism, etc. The Clavien-Dindo system grades the severity of complications based on the impact of the complication on the management of the patient and has been validated in large numbers of surgical patients.\parencite{clavien_clavien-dindo_2009, dindo_classification_2004} The classification is summarised in Table \ref{table:clavien-dindo} on p\pageref{table:clavien-dindo} and was used to grade complications in this thesis. Complications of grade 3 or greater are considered clinically significant as they require further procedures either under local or general anaesthesia, single or multiple organ support or result in death while grades 1 and 2 are managed with ward-based care including antibiotics, wound management, supplemental enteral or parenteral nutrition and similar interventions.

@@

While surgery is the primary modality of curative treatment in patients with resectable pancreatic cancer, there is clear evidence that adjuvant chemotherapy improves survival and is now the standard of care for patients with pancreatic ductal adenocarcinoma. The European Study Group for Pancreatic Cancer (ESPAC) has conducted some of the largest randomised controlled trials of adjuvant chemotherapy and chemoradiotherapy versus observation after potentially curative surgery and have reported that chemotherapy prolonged survival after surgery while chemoradiotherapy had a harmful effect. \parencite{neoptolemos_adjuvant_2001, neoptolemos_randomized_2004, neoptolemos_adjuvant_2009, neoptolemos_adjuvant_2010}

The ESPAC-1 trial compared the use of chemotherapy with Fluorouracil and Folinic acid or chemoradiotherapy with 20Gy radiotherapy and 5-Fluorouracil (5-FU) versus observation only in patients who underwent potentially curative surgery for pancreatic ductal adenocarcinoma. \parencite{neoptolemos_randomized_2004} The 5-year survival rate in patients who received chemotherapy was 20\% while it was 8\% in patients who did not receive chemotherapy, demonstrating a clear survival benefit for adjuvant chemotherapy. However, adjuvant chemoradiotherapy resulted in 5-year survival dropping from 20\% to 10\% and has largely been abandoned in the adjuvant setting. The ESPAC-3 trial compared the efficacy of 5-FU and Folinic acid with Gemcitabine and reported that Gemcitabine was not superior to 5-FU/Folinic acid. \parencite{neoptolemos_adjuvant_2010}

However, 25\% to 50\% of patients who undergo surgery for pancreatic ductal adenocarcinoma either do not receive or do not complete a full course of adjuvant chemotherapy. \parencite{sohn_resected_2000, neoptolemos_randomized_2004, oettle_adjuvant_2007} A number of factors have been identified for the non-receipt of adjuvant therapy including comorbidity, postoperative complications and favourable pathological features.\parencite{spitz_preoperative_1997, aloia_delayed_2007, oettle_adjuvant_2007, merchant_adjuvant_2009, russ_impact_2009}

@@

More recently, cross-sectional imaging including computed tomography (CT) and magnetic resonance imaging (MRI) have become widely available. Nearly all patients with gastrointestinal cancer undergo either CT or MRI or both before commencing treatment. Accurate measurements of subcutaneous adipose tissue, visceral adipose tissue and skeletal muscle can be made relatively easy either from the entire series of images or estimates can be made from measurements made at certain predetermined anatomical levels. Both specialised software or readily available generic image analysis tools have been used to make these measurements. The proportion of the total cross-sectional area that is made of adipose tissue or skeletal muscle at the level of the 3$^{rd}$  or the 4$^{th}$ lumbar vertebra is commonly used as a surrogate for total body measurements. Thickness of retro-renal visceral fat at the level of the renal vein has also been reported to be usable as a surrogate for total visceral adiposity. Studies have also reported on the association between intra-pancreatic fat content and postoperative complications.
